---
title: "Analysis of Exponential Distribution and Central Limit Theorem"
author: "Kevin E. D'Elia"
date: "January 25, 2016"
output: html_document
---

File needs to be in PDF format.
Verify that you can open it from the Coursera submission page.

Does the title adequately summarize the analysis?

# Synopsis of Analysis
describe and summarize the data analysis in a few sentences (2-3)

There are two parts to this project:
1. A simulation exercise to explore inference
2. Basic inferential data analysis


This is a good reference, and you can include it and cite portions of it.
https://en.wikipedia.org/wiki/Exponential_distribution


The exponential distribution can be simulated in R with **rexp(n, lambda)** where $\lambda$ is the *rate* parameter.  The mean of exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also $\frac{1}{\lambda}$.  A requirement of this report is to set $\lambda$ = 0.2 for all of the simulations.

Investigate the distribution of averages (means) of 40 exponentials (n = 40). 
so run rexp in a loop 1000 times and collect the results.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials.  You should:
1. show the sample mean and compare it to the theoretical mean of the distribution
2. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
3. Show that the distribution is approximately normal (How to do this???)
  for point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.
  An example using random uniforms (replace with random exponentials) is this:
  
  # distribution of 1000 random uniforms
```{r}
hist(runif(1000))
```

# distribution of 1000 averages of 40 random uniforms
```{r}
mns <- NULL
for (i in 1:1000) mns <- c(mns, mean(runif(40)))
hist(mns)
```

Do not forget to set your seed!!!

Add this ->

By The Central Limit Theorem, for large n, $\bar{X}$ ~ N($\mu$, $\sigma^2$/n)

# Report Structure
USE THIS AS YOUR STARTING POINT AND WORK THE REST IN!!
## Title and Author name 
(already in yaml)
## Overview 
what is going to be reported on in 2-3 sentences
## Simulations 
which simulations were run, with R code.  Explanations should make clear what the R code accomplishes
## Sample Mean versus Theoretical Mean: 
include figures with titles.  In the figures, highlight the means you are comparing.  Include ext that explains the figures and what is shown on them and provides appropriate numbers
## Sample Variance versus Theoretical Variance:  
include figures (output from R) with titles.  Highlight the variances you are comparing.  Include text that explains your understanding of the differences of the variances.
## Distribution 
via figures and text, explain how one can tell the distribution is approximately normal.

Make sure you show the center of the distribution of the mean of the 40 exponentials and compare it to the theoretical center of the distribution.
Make sure you show the variance and compare it to the theoretical variance of the distribution.

Need explanations of the simulations run, along with the accompanying R code

The document compares the sample mean to the theoretical mean, including a figure (table or graph) with a relevant title.  The document shows where the distribution is centered, and compares it to the theoretical center of the distribution.

The document compares the sample variance to the theoretical variance, including a figure (table or graph) with a relevant title.  The document shows the variability of the mean of 40 exponentials, and compares it to the theoretical variance /standard error of the distribution.

The document describes in figures(s) and text how one can tell whether the distribution is normal.

The document answers the question "Does the distribution of means of 40 exponentials behave as predicted by the Central Limit theorem?"

The Report section before the Appendix is <= 3 pages
The Appendix is 3 pages or fewer and separates the report from supporting code and figures

Document has been proof read and spell-checked

Coding statements do not extend beyond the right margin


EXAMPLE:

to generate 100 random # from an exponential distribution w/mu = 50
vector <- rexp(100, 50)
where 50 is the rate param and mean is 1/rate
Use a histogram to plot this


```{r}
# Dummy values
theMean = 2.1589
theSd = 3.989388
Statistic <- c("Mean", "Std. Deviation")
Sample <- c(round(theMean, 2), round(theSd, 2))
Theoretical <- c(5.0, 5.0)
theTable <- data.frame(Statistic, Sample, Theoretical)
rownames(theTable) <- NULL
knitr::kable(theTable)
```
