---
title: "Analysis of Exponential Distribution and Central Limit Theorem"
author: "Kevin E. D'Elia"
date: "January 25, 2016"
output: pdf_document
---

# Overview (Synopsis of Analysis)
The Central Limit Theorem (CLT), and the concept of the sampling distribution, are critical for understanding why statistical inference works.  The CLT says that if you take many repeated samples from a population, and calculate the averages (or sum) of each one, the collection of those averages will be normally distributed.  The purpose of this project is to examine the mean and variance of a sample from the exponential distribution, then produce a sampling distribution of the mean and variance, and then demonstrate how the sampling distribution is approximately normal.

# Simulations

As with any good simulation, the first step is to set the seed for the random number generators.  This is done so that the results generated by the random number engine will be consistent across all invocations of the code, thus ensuring that the results are reproducible by other interested parties.

```{r}
set.seed(1234)
```

The R code for the simulations will be included in-line with the report for easy verification, along with explanations about what is being accomplished by each R markdown code chunk.

There will be two simulations run.  The first will draw 40 values from the exponential distribution.  The second will perform the first simulation, except that it will be executed 1000 times and, for each execution, the mean and variance will calculated for the sample and stored in a vector. 

The exponential distribution can be simulated in R with **rexp(n, lambda)** where $\lambda$ is the *rate* parameter.  The mean of exponential distribution is $\frac{1}{\lambda}$ and the standard deviation is also $\frac{1}{\lambda}$.  A requirement of this report is to set $\lambda$ = 0.2 for all of the simulations.

I chose 10,000 replications because this number is large enough to ensure that the mean of the sampling distribution approaches the mean of the population.

Investigate the distribution of averages (means) of 40 exponentials (n = 40). 
so run rexp in a loop 1000 times and collect the results.

# Sample Mean versus Theoretical Mean: 

include figures with titles.  In the figures, highlight the means you are comparing.  Include ext that explains the figures and what is shown on them and provides appropriate numbers

show the sample mean and compare it to the theoretical mean of the distribution

# Sample Variance versus Theoretical Variance:  

include figures (output from R) with titles.  Highlight the variances you are comparing.  Include text that explains your understanding of the differences of the variances.

Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.

# Distribution 

By The Central Limit Theorem, for large n, $\bar{X}$ ~ N($\mu$, $\sigma^2$/n)

via figures and text, explain how one can tell the distribution is approximately normal.
Show that the distribution is approximately normal (How to do this???)
  for point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.
  An example using random uniforms (replace with random exponentials) is this:
  
  # distribution of 1000 random uniforms
```{r}
hist(runif(1000))
```

# distribution of 1000 averages of 40 random uniforms
```{r}
mns <- NULL
for (i in 1:1000) mns <- c(mns, mean(runif(40)))
hist(mns)
```



Make sure you show the center of the distribution of the mean of the 40 exponentials and compare it to the theoretical center of the distribution.
Make sure you show the variance and compare it to the theoretical variance of the distribution.

Need explanations of the simulations run, along with the accompanying R code

The document compares the sample mean to the theoretical mean, including a figure (table or graph) with a relevant title.  The document shows where the distribution is centered, and compares it to the theoretical center of the distribution.

The document compares the sample variance to the theoretical variance, including a figure (table or graph) with a relevant title.  The document shows the variability of the mean of 40 exponentials, and compares it to the theoretical variance /standard error of the distribution.

The document describes in figures(s) and text how one can tell whether the distribution is normal.

The document answers the question "Does the distribution of means of 40 exponentials behave as predicted by the Central Limit theorem?"

The Report section before the Appendix is <= 3 pages
The Appendix is 3 pages or fewer and separates the report from supporting code and figures

Document has been proof read and spell-checked

Coding statements do not extend beyond the right margin


EXAMPLE:

to generate 100 random # from an exponential distribution w/mu = 50
vector <- rexp(100, 50)
where 50 is the rate param and mean is 1/rate
Use a histogram to plot this


```{r}
# Dummy values
theMean = 2.1589
theSd = 3.989388
Statistic <- c("Mean", "Std. Deviation")
Sample <- c(round(theMean, 2), round(theSd, 2))
Theoretical <- c(5.0, 5.0)
theTable <- data.frame(Statistic, Sample, Theoretical)
rownames(theTable) <- NULL
knitr::kable(theTable)
```
