---
title: "Analysis of Guinea Pig Tooth Growth Using R"
author: "Kevin E. D'Elia"
date: "March 5, 2016"
output: pdf_document
---

# Synopsis of Analysis
In this paper, the techniques of Exploratory Data Analysis (EDA) will be applied to the dataset __ToothGrowth__.  Among those techniques are the use of tabular (_summary()_, _str()_) as well as graphical (_histogram()_, _boxplot()_) representations of the data in order to glean patterns and structure within the data.  The analysis will attempt to answer the following question:  Is the data in this dataset normally distributed and, by extension, if not, then why not?

# Summary of the data
Usually the first step in any analysis is to acquire and load the data.  In this case, however, the dataset is preloaded in __R__ and is available via the __datasets__ package.  So, simply typing __ToothGrowth__ at the console prompt will display the full dataset.  The first few lines look like this:
```{r display_head}
head(ToothGrowth)
```

Before doing any summary analysis of the data, it is helpful to examine its structure, which is accomplished like so:
```{r structure}
str(ToothGrowth)
```

This table indicates that the dataset consists of 60 observations of 3 variables.  Since the second variable is a factor, or grouping, variable, the next step in the analysis will exclude it and look only at numerical data.  To see a basic summary of the ToothGrowth numerical data, use the eponymous R command:
```{r summarize}
summary(ToothGrowth[c(1, 3)])
```

provide a basic summary

```{r}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth, xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

### ---- Next area of focus
Document outlines key assumptions made about the analysis such as whether the variables are normally distributed, and how this affects the choice of statistical methods used for the analysis

### --- Confidence intervals
document uses confidence intervals and/or hypothesis tests to compare tooth growth by dose and supp (Delivery Method)

##  -- Results
Results of the confidence intervals / hypothesis tests are interpreted correctly, given the context of the problem.

## -- Answers
document provides an answer to the question, "Do Delivery methods and/or dosage affect tooth growth in guinea pigs?"

The Report section before the Appendix is <= 3 pages
The Appendix is 3 pages or fewer and separates the report from supporting code and figures

Document has been proof read and spell-checked

Coding statements do not extend beyond the right margin


